Smart Spreading Factor Assignment for LoRaWANs

### 2019-01-18 paper.tex
advisor advives

\begin{abstract}
Low power wide area network (LPWAN) technologies offer affordable connectivity to massive number of low-power devices distributed over large geographical areas. Focus of this work is one of the most promising LPWAN technologies: LoRa. LoRa offers long range communication and strong resilience to interference by proprietary modulation technique based on Chirp Spread Spectrum (CSS). LoRa modulation trades data rate for sensitivity and communication range by spreading symbols within a fixed channel bandwidth. Collisions in LoRaWAN networks are strongly related with spreading factor (SF) assignment of nodes which indeed effects network performance. In this work, a simulation environment to study different SF assignment schemes is implemented. Furthermore, a novel SF assignment strategy is proposed which utilizes a machine learning technique for optimization of SF assignment. Finally, promising simulation results for proposed machine learning SF assignment technique is presented.
\end{abstract}

@@

\par There are several emerging LPWAN technologies. LoRa, Sigfox, NB-IoT and LTE-M are commonly used, well-known LPWAN technologies. LoRa and Sigfox use license free ISM frequency bands while NB-IoT and LTE-M use licensed frequency bands which brings extra cost \cite{7815384}. Both LoRa and Sigfox are known for ultra-low power consumption and resilience to interference. While NB-IoT and LTE-M are promoted for higher data rate. LoRa has open standard MAC protocol called LoRaWAN. LoRaWAN and Sigfox MAC protocols are based on pure ALOHA medium access. LoRaWAN networks can be deployed as a private network like WiFi. However, Sigfox and NB-IoT are only available with operator contract \cite{7815384}.

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and time on air of transmission \cite{7803607}. Simultaneous same SF transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other, thus SF assignment is crucial for overall network performance. In this work, a LoRa discrete event simulator is developed to study different LoRa SF assignment strategies. Also an SF assignment scheme is proposed to increase network performance with the help of machine learning methods and the proposal is verified using the simulator.

\par The following two sections provides a background information about LoRa and LoRaWAN. Section \ref{Other Related Works} describes other related works. Section \ref{Proposed Technique} describes proposed technique for SF assignment. Simulation environments and results are shown in Section \ref{Simulation Environment} and \ref{Simulation Results}. Finally, Section \ref{Conclusion} concludes the paper.

\section{LoRa}
\par LoRa is the name of the proprietary physical layer radio/chipset technology that provides wireless link solution for LPWANs. LoRa is designed and patented by Semtech Corporation. LoRa uses proprietary spread spectrum modulation technique that is derivative of Chirp Spread Spectrum (CSS). A chirp is a sinusoidal signal which its frequency increases over time. Chirp frequency increases linearly and sweeps the entire bandwidth. We invite readers to refer to \cite{AN1200.22} for further details.

\subsection{Spreading Factor}
\par Spreading factor (SF) is the ratio between symbol rate and chirp rate. The ratio between symbol and chirp rate is equal to $2$\textsuperscript{SF}. SF can take values between 7 to 12. SF also determines data rate of a LoRa transmission \cite{AN1200.22}. Data rate of a LoRa transmission can be calculated as:

@@

\par In Figure \ref{fig:collision}, a LoRaWAN network deployed with a single GW is illustrated. Different color rings represent achievable range of different SFs from the GW and different color circles represent selected SF of the nodes. Most of the end devices close to the GW will fall into lowest SF (SF7) area section. End devices close to the GW will probably select lowest SF all the time. This causes a lot of collisions between same SF transmissions. Number of collisions increases while number of end devices close to the GW increases.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa has an open standard medium access control (MAC) layer protocol called LoRaWAN is which designed for large scale LoRa networks considering well known LPWAN challenges and their best practice solutions. LoRaWAN is developed and maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRaWAN is based on pure ALOHA medium access which means that end nodes do not check whether the channel is free or not before transmission, accepting the possibility of a collision. A typical LoRaWAN network consists of following three network entities.

\subsection{End node}
\par LoRaWAN end node (EN) is a low power embedded device that only communicates to GWs. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C. Different classes provide LPWAN solutions to different applications and deployments. Class A end nodes generate uplink transmission at any time and only receive a period of time after uplink transmission. Class B end nodes extend Class A behavior by adding scheduled receive windows for downlink transmission. Receive window is synchronized using a beacon packet transmitted by GWs. Class C end nodes extend Class A behavior by keeping receive window open all the time except uplink transmission. This provides Class C end nodes with low latency downlink communication, but it requires more power consumption. In this paper, only Class A end devices are considered since Class A behavior provides the lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway (GW) is a device that communicates to end nodes. A typical GW can receive from multiple channels at the same time. GWs are usually connected to power grid, so power consumption of a GW is insignificant in most of the deployments.

@@

\section{Other Related Works} \label{Other Related Works}
\par The literature related to the work presented in this paper has started growing recently. LPWAN technologies and especially LoRa attracted researchers’ attentions lately. Some of these works which studies LoRa/LoRaWAN SF are listed.

@@

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN Class A end devices, multiple GWs, both upstream and downstream confirmed messages. Their results show that allocating network parameters to end devices is highly important for the performance of LoRaWAN networks.

\par In \cite{8267219}, the authors studied imperfect orthogonality between different LoRa SF transmissions. The authors state that a LoRa transmission can be interfered even between different SF transmissions when power of the interfering signal significantly overcomes the reference signal. Their experimental results show that this power difference is around 16 dB and this power difference can be seen when an interfering signal is close to a receiver or sum of interfering signals' energy can create this power difference.

\par In \cite{8430542}, the authors investigated the impact of interference caused by simultaneous LoRa transmissions with the same and different SFs. They derived aggregated co-SF and inter-SF interference power SIR distributions to capture the coverage distance from the GW for modeling interference in multiple GWs scenarios. Their results show that transmission among different SFs can cause a significant impact in high-density LoRaWAN networks.

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} is solved by forcing some of the close nodes to select higher SF even they are able to communicate with lower SF. This has potential to prevent collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, distribution of SF among nodes becomes an important question. Increasing a node's SF should be done carefully since higher SF means longer time on air and longer time on air is more prone to collision with other high SF transmissions. In multiple GWs scenarios, this approach may increase the collisions with nodes in other GW's range. Thus, extra care should be taken for nodes in intersection area of the GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes' SFs requires different rules. For this reason, machine learning based SF assignment approach is proposed to decrease same SF transmission collisions. This technique works by learning the transmission behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet is not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than sensitivity of the GW. Using these obtained information, NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the SF to nodes considering the collision probability.

\par In this work, decision tree classifier (DTC) and support vector machine (SVM) schemes are evaluated to predict transmission results. Class weights are balanced according to sample distributions for both methods. For DTC, Gini impurity criteria is used to measure the quality of a splits. For SVM, penalty parameter is set to 1, degree is set to 3 and RBF kernel is used.

\par It is possible to generate mass amount of LoRaWAN transmission records for different topologies using the simulator. Data set size is directly proportional to simulation duration. Thus, increasing simulation duration, improves the prediction accuracy. In real world deployments, NS can keep track of all the transmission and it can create a classifier daily basis, then GWs can request from nodes to use their new SFs.

\par To come up with a solution for SF assignment issue, DTC and SCM prediction schemes are integrated to the simulator. A classifier is trained from transmission logs and this classifier is used for selecting optimum SF for the nodes. Simulator first runs a simulation with assigning random SF to the nodes. Simulation results are combined into four data columns which are; X and Y coordinates of the transmission source, SF of the transmission and result of the transmission. Possible transmission results are successful, interfered or under sensitivity. These data are fed to Python scikit-learn DTC or SVM classifiers. After the classifier model is generated, a second simulation is run with classifier. Simulator selects optimum SF considering prediction of transmission. If a transmission of a node with lowest possible SF is predicted as interfered by the classifier, then simulator assigns higher SF and calculates new prediction. If no SF is predicted as successful, then simulator again assigns the lowest possible SF.

@@

\section{Simulation Environment} \label{Simulation Environment}
\par A discrete event simulator is implemented in Python to study effects of different SF strategies in LoRaWANs. LoRaWAN SF simulation tool source code is available at \cite{simlorasf}. Simulation tool supports custom LoRaWAN topologies as well as randomly generated LoRaWAN topologies. Simulator can generate uniformly distributed circular shape network topology with input parameters; radius (m), number of nodes and number of GWs. Global simulation input parameters are simulation duration (s), packet size (B), packet generation rate (p/s) and SF assignment method. With these inputs, the simulator produces total number of generated packets, number of successfully received packets, number of interfered packets, number of under sensitivity packets, network packet delivery ratio percentage (PDR), network throughput (bps) and total transmit energy consumption (J). Simulator also produces prediction accuracy and confusion matrix for machine learning schemes.

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet randomly according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictate duty cycle limit for all devices including GWs.

\subsection{Link Model}
\par Link quality of a wireless system can be expressed by the metric of link budget. Link budget is a measure of all gains and losses from transmitter device to receiver device. Link budget of a wireless link can be calculated as \cite{AN1200.22}:

@@

\subsection{Interference Model}
\par In the simulator, interference model described in \cite{7996384} is adopted. They use SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions.

\par In our simulator, it is assumed that there is no other technology interference in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, simulator should calculate the effect of different SF transmissions to each other. In simulator, signal to interference plus noise ratio (SINR) threshold matrix from \cite{goursaud:hal-01231221} is used.

\par To decide if a referenced signal is interfered at receiver by an interfering signal, SINR threshold matrix is used. $T_{i,j}$ is the needed SINR margin in dB unit between referenced signal with SF = i and interfering signal with SF = j to correctly decode the referenced signal. If there are more than one interfering signal, referenced signal must satisfy the margin for cumulative sum of all interfering signal received power for each SF \cite{7996384}. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}.

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, global simulation parameters are set as follows; packet size is 60 bytes including header and payload, simulation duration is 3600 seconds, packet generation rate is 0.01 packet per second.

\par In Figure \ref{fig:sf_pdr}, PDR plot of different SF assignment is shown. Randomly generated network topology radius is set to 3000 meters and number of GWs is set to 1. Increasing SFs, increases time on air. This increases the number of collisions thus decreases the PDR. High SF schemes gives poor PDR results while number of nodes increases. Since network topology radius is quite short, all SFs can reach to the GW.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{sf_pdr_r3000_g1_p0.01_s3600}.png}
\caption{PDR of different SFs.}
\label{fig:sf_pdr}
\end{figure}

\par In Figure \ref{fig:gw_pdr}, PDR plot of different number of GWs is shown. Randomly generated topology radius is set to 3000 meters and lowest possible SF assignment scheme is used. Increasing number of GWs, decreases the SFs of nodes and decreases time on air. This decreases number of collisions thus increases the PDR of the network when network topology radius is constant.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{gw_pdr_r3000_p0.01_s3600}.png}
\caption{PDR of different number of GWs.}
\label{fig:gw_pdr}
\end{figure}

\par In Figure \ref{fig:r_pdr}, PDR plot of different network topology radius is shown. Number of GW is set to 1 and lowest possible SF assignment scheme is used. Increasing network topology radius, increases the number of under sensitivity transmissions thus decreases the PDR of the network.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{r_pdr_g1_p0.01_s3600}.png}
\caption{PDR of different topology radius.}
\label{fig:r_pdr}
\end{figure}

\par For space constraints, we omit figures for network throughput output, total transmit energy consumption output and different packet generation rate input. We invite readers to experiment the simulation tool with different input parameters and their results. \cite{simlorasf}.

\par In Figure \ref{fig:prediction_pdr}, result of proposed technique is presented. PDR plot of lowest SF scheme and machine learning prediction schemes are shown in the same figure. Randomly generated network topology radius is set to 5000 meters and number of GWs is set to 3. Prediction model needs nodes location and three GWs are enough to locate nodes positions by triangulation. In Table \ref{table:prediction_pdr}, PDR values for various topologies with different radii are presented. Both prediction SVM and DTC SF schemes give better PDR than lowest SF schemes when number of nodes increases. Increasing number of nodes, increases number of interference. Prediction schemes  improve network performance when LoRa interference is high. Also, prediction schemes give better results when nodes are deployed close to the GW, since nodes have margin to increase their SFs when they are deployed close to the GW. If a node is far away from the GW, then prediction schemes cannot increase the SF to avoid interference since the SF is already high.

\par In Table \ref{table:prediction_accuracy}, prediction accuracy of SVM and DTC schemes are presented for different network topology radius and number of nodes. Prediction accuracy is not directly proportional to network PDR. Correct prediction of a interfered transmission may not increase the PDR but increases the accuracy metric. DTC gives better network PDR results than SVM, even overall SVM prediction accuracy is high. LoRaWAN simulation usually produces imbalanced dataset, DTC gives better accuracy for interfered transmission labels. We invite readers to experiment the simulation tool to review confusion matrices for SVM and DTC schemes.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{prediction_pdr_r5000_g3_p0.01_s3600}.png}
\caption{PDR of SF prediction scheme.}
\label{fig:prediction_pdr}
\end{figure}

\begin{table}
\centering
\caption{PDR of lowest possible SF and smart SF assignment schemes.}
\label{table:prediction_pdr}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{\multirow{2}{*}{}}                            & \multicolumn{3}{c|}{\textbf{Number of Nodes}} \\ \cline{4-6}
\multicolumn{3}{|c|}{}                                             & 100           & 500           & 1000          \\ \hline
\multirow{12}{*}{\textbf{R (m)}} & \multirow{3}{*}{3000}  & Lowest & 97.8          & 86.0          & 72.3          \\ \cline{3-6}
                                 &                        & SVM    & 98.0          & 88.2          & 75.2          \\ \cline{3-6}
                                 &                        & DTC    & 97.8          & 89.8          & 78.7          \\ \cline{2-6}

@@

\section{Conclusion} \label{Conclusion}
\par In this work, after a brief introduction about LPWAN technologies, background information about LoRa and LoRaWAN is provided. Then, LoRa modulation basics and spreading factor assignment issue is  discussed. A discrete event simulator is implemented from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. Same SF transmission collisions are shown and machine learning based solution is proposed for collision with same SF transmissions. Simulation results for proposed machine learning SF assignment technique is illustrated. Simulation results show that, proposed technique can increase network performance for LoRaWAN networks especially nodes are deployed close to the GWs.

\par As for future works, transmit power optimization can be included in proposed machine learning scheme. In this paper, it is assumed that nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmissions more vulnerable to interference thus requires extra care. Also, other machine learning methods can be investigated for SF optimization. Reinforcement learning is a good candidate for this network optimization issue. Other transmission parameters such as node id and transmission time can be included to proposed scheme to improve performance.

>>>>

\begin{abstract}
Low power wide area network (LPWAN) technologies offer affordable connectivity to massive number of low-power devices distributed over large geographical areas. Focus of this work is one of the most promising LPWAN technologies: LoRa. LoRa offers long range communication and strong resilience to interference by proprietary modulation technique based on Chirp Spread Spectrum (CSS). LoRa modulation trades data rate for sensitivity and communication range by spreading symbols within a fixed channel bandwidth. Collisions in LoRaWAN networks are strongly related with spreading factor (SF) assignment of nodes which indeed effects network performance. In this work, a simulation environment to evaluate the performance of SF assignment schemes is implemented. Furthermore, a novel smart SF assignment strategy which utilizes Support Vector Machine (SVM) and Decision Tree Classifier (DTC) machine learning techniques for optimization of SF assignment is proposed. It is observed and presented that the proposed smart SF assignment techniques give promising simulation results. 
\end{abstract}

@@

\par There are several emerging LPWAN technologies. LoRa, Sigfox, NB-IoT and LTE-M are commonly used, well-known LPWAN technologies \cite{7815384}. LoRa and Sigfox use license free ISM frequency bands while NB-IoT and LTE-M use licensed frequency bands which brings extra cost \cite{7815384}. Both LoRa and Sigfox are known for ultra-low power consumption and resilience to interference. While NB-IoT and LTE-M are promoted for higher data rate. LoRa has open standard MAC protocol called LoRaWAN. LoRaWAN and Sigfox MAC protocols are based on pure ALOHA medium access. LoRaWAN networks can be deployed as a private network like WiFi. However, Sigfox and NB-IoT are only available with operator contract \cite{7815384}.

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and air time of transmission \cite{7803607}. Simultaneous same SF transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other. Thus SF assignment is crucial for overall network performance. In this work, a LoRa discrete event simulator is developed to study the performance of LoRa SF assignment strategies. Also a machine learning based SF assignment approach is proposed. Support Vector Machine (SVM) and Decision Tree Classifier (DTC) techniques are employed and the introduced schemes are called as smart SF schemes. The performance of the smart schemes  is compared with the performance of the lowest SF assignment scheme. It is shown that the proposed smart SF assignment schemes give promising results.

\par This paper is organized as follows: Section \ref{LoRa} and Section \ref{LoRaWAN} provides some background information about LoRa and LoRaWAN. Section \ref{Related Works} summarize  the other related works. Section \ref{Proposed Technique} describes the smart SF assignment technique proposed. Simulation environments and results are shown in Section \ref{Simulation Environment} and \ref{Simulation Results}. Finally, Section \ref{Conclusion} concludes the paper by givin future directions.

\section{LoRa} \label{LoRa}
\par LoRa is a proprietary physical layer radio/chipset technology that provides wireless link solution for LPWANs. It uses proprietary spread spectrum modulation technique that is the derivative of Chirp Spread Spectrum (CSS). A chirp is a sinusoidal signal of which frequency increases over time. Chirp frequency increases linearly and sweeps the entire bandwidth \cite{AN1200.22}.

\subsection{Spreading Factor}
\par The ratio between symbol and chirp rate is equal to $2$\textsuperscript{SF}. SF can take values between 7 to 12. SF also determines data rate of a LoRa transmission \cite{AN1200.22}. Data rate of a LoRa transmission can be calculated as:

@@

\par In Figure \ref{fig:collision}, a LoRaWAN network deployed with a single GW is illustrated. Different color rings represent achievable range of different SFs from the GW and different color circles represent selected SF of the nodes. The end devices close to the GW will fall into lowest SF (SF7) area section. The end devices close to the GW will probably select the lowest SF mot the time.Hence the number of collisions will increase as the number of end devices close to the GW increases.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa has an open standard medium access control (MAC) layer protocol called LoRaWAN is which is designed for large scale LoRa networks considering well known LPWAN challenges and their best practice solutions. LoRaWAN is developed and maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRaWAN is based on pure ALOHA medium access which means that end nodes do not check whether the channel is free or not before transmission, accepting the possibility of a collision. A typical LoRaWAN network consists of following three network entities.

\subsection{End node}
\par LoRaWAN end node (EN) is a low power embedded device that only communicates to GWs. LoRaWAN standard defines three classes for end devices which are Class A, Class B, and Class C. Different classes provide LPWAN solutions to different applications and deployments. Class A end nodes generate uplink transmission at any time and only receive a period of time after uplink transmission. Class B end nodes extend Class A behavior by adding scheduled receive windows for downlink transmission. Receive window is synchronized using a beacon packet transmitted by GWs. Class C end nodes extend Class A behavior by keeping receive window open all the time except uplink transmission. This provides Class C end nodes with low latency downlink communication, which it requires more power consumption. In this paper, only Class A end devices are considered since Class A behavior leads to the lowest power consumption.

\subsection{Gateway}
\par LoRaWAN gateway (GW) is a device that communicates with the end nodes. A typical GW can receive from multiple channels at the same time. GWs are usually connected to power grid, so power consumption of a GW is insignificant in most of the deployments.

@@

\section{Related Works} \label{Related Works}
\par The literature related to the work presented in this paper has started to grow recently. LPWAN technologies and especially LoRa attracted researchers’ attentions lately. Some of these works which studies LoRa/LoRaWAN SF are summarized.

@@

\par In \cite{8090518}, another LoRaWAN ns-3 simulator is presented. Authors introduced an error model for determining range as well as interference between multiple simultaneous LoRa transmissions. Their simulator supports LoRaWAN Class A end devices, multiple GWs, both upstream and downstream confirmed messages. Their results show that allocating network parameters such as SF is highly important for the performance of LoRaWAN networks.

\par In \cite{8267219}, the authors studied the effects of imperfect orthogonality between different LoRa SF transmissions. The authors state that a LoRa transmission can be interfered by a different SF transmission when power of the interfering signal significantly is greater than the reference signal. Their experimental results show that this power difference is around 16 dB such a this power difference can be seen when an interfering signal is close to a receiver or the sum of interfering signals' energy can create this power difference.

\par In \cite{8430542}, the authors investigated the impact of the interference caused by simultaneous LoRa transmissions with the same and different SFs. They derived aggregated co-SF and inter-SF interference power SIR distributions to capture the coverage distance from the GW for modeling interference in multiple GWs scenarios. Their results show that transmission among different SFs can cause a significant impact in high-density LoRaWAN networks.

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} is solved by forcing some of the close nodes to select higher SFs even they are able to communicate with lower SFs. This has potential to prevent collisions due to the orthogonality of the SFs as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, distribution of SF among nodes becomes an important problem. Increasing a node's SF should be done carefully since higher SF means longer air time and longer air time means increasing the probability of collisions with other high SF transmissions. In multiple GWs scenarios, this approach may increase the collisions with the nodes in other GW's range. Thus, extra care should be taken for nodes in the intersection area of the GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes' SFs requires different rules. For this reason, machine learning based SF assignment approach is proposed to decrease the colisioins for the same SF transmissions. This technique starts by learning the transmission behavior of the nodes in a network. An NS can keep track of successful uplink transmissions and their SFs. This NS can also keep track of some of the collided transmissions if header part of the packet is not interfered at the GW. However, NS cannot keep track of transmissions with lower receive power than sensitivity of the GW. Using this information, NS can train a classifier to predict future transmission result for a specific node and a specific SF. Using this prediction model NS can assign SFs to nodes considering the collision probability.

\par In this work, decision tree classifier (DTC) and support vector machine (SVM) \cite{Alpaydin} schemes are employed to predict the transmission results. Class weights are balanced according to sample distributions for both methods. For DTC, Gini impurity criteria is used to measure the quality of splits. For SVM, penalty parameter is set to 1, degree is set to 3 and RBF kernel is used.

\par It is possible to generate mass amount of LoRaWAN transmission records for different topologies using our simulator. Definetely, data set size is directly proportional to simulation duration. Thus, increasing the simulation duration, improves the prediction accuracy. In real world deployments, NS can keep track of all transmissions and it can create a classifier daily basis, then GWs can request from nodes to use suggested SFs.

\par To come up with a solution for SF assignment, DTC and SVM prediction schemes are integrated to the simulator. A classifier is trained from transmission logs and this classifier is used for selecting optimum SF for the nodes. Simulator first runs a simulation by assigning random SFs to the nodes. Simulation logs are combined into four data columns which are; X and Y coordinates of the transmission source, SF of the transmission and the result of the transmission. Possible transmission results are successful, interfered or under sensitivity. These data are fed to Python scikit-learn DTC or SVM classifiers. After the classifier model is generated, a second simulation is run with classifier. Simulator selects optimum SF considering the prediction of transmission. If a transmission of a node with the lowest possible SF is predicted as interfered by the classifier, then simulator assigns higher SF and evaluates the new prediction. If no SF is predicted as successful, then simulator again assigns the lowest possible SF.

@@

\section{Simulation Environment} \label{Simulation Environment}
\par A discrete event simulator is implemented in Python to study the effects of different SF strategies in LoRaWANs. LoRaWAN SF simulation tool source code is available at \cite{simlorasf}. Simulation tool supports custom LoRaWAN topologies as well as randomly generated LoRaWAN topologies. Simulator can generate uniformly distributed circular shape network topology with input parameters; radius (m), number of nodes and number of GWs. Global simulation input parameters are simulation duration (s), packet size (B), packet generation rate (p/s) and SF assignment method. With these inputs, the simulator produces total number of generated packets, number of successfully received packets, number of interfered packets, number of under sensitivity packets, network packet delivery ratio percentage (PDR), network throughput (bps) and total transmit energy consumption (J). Simulator also produces prediction accuracy and confusion matrix for machine learning schemes.

\par The simulator only covers the LoRaWAN Class A devices. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet randomly according to a Poisson interval for given packet rate parameter. Downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictate duty cycle limit for all devices including GWs.

\subsection{Link Model Employed}
\par Link quality of a wireless system can be expressed by the metric of link budget. Link budget is a measure of all gains and losses from transmitter device to receiver device. Link budget of a wireless link can be calculated as \cite{AN1200.22}:

@@

\subsection{Interference Model Employed}
\par In the simulator, interference model described in \cite{7996384} is adopted. They use SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions.

\par It is assumed that there is no other technology interference in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, simulator should calculate the effect of different SF transmissions to each other. In simulator, signal to interference plus noise ratio (SINR) threshold matrix from \cite{goursaud:hal-01231221} is used.

\par To decide if a referenced signal is interfered at receiver by an interfering signal, SINR threshold matrix is used. $T_{i,j}$ is SINR margin in dB unit between referenced signal with SF = i and interfering signal with SF = j to correctly decode the referenced signal. If there are more than one interfering signal, referenced signal must satisfy the margin for cumulative sum of all interfering signal received power for each SF \cite{7996384}. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}.

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, global simulation parameters are set as follows: packet size = 60 Bytes, simulation duration = 3600 sec, packet generation rate = 0.01 pac/sec.

\par In Figure \ref{fig:sf_pdr}, PDR plot of various SF assignment is shown. Randomly generated network topology radius is set to 3000 meters and number of GWs is set to 1. Increasing SFs, increases air time. It increases the number of collisions thus decreases the PDR. High SF schemes gives poor PDR results as the number of nodes increases. Since network topology radius is quite small, all SFs can reach to GW.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{sf_pdr_r3000_g1_p0.01_s3600}.png}
\caption{PDR for various SFs.}
\label{fig:sf_pdr}
\end{figure}

\par In Figure \ref{fig:gw_pdr}, PDR plot for various number of GWs is shown. Randomly generated topology radius is set to 3000 meters and the lowest SF assignment scheme is used. Increasing number of GWs, decreases the SFs of nodes and decreases air time. This decreases number of collisions, henc, increases the PDR of network when network radius is constant.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{gw_pdr_r3000_p0.01_s3600}.png}
\caption{PDR for various number of GWs.}
\label{fig:gw_pdr}
\end{figure}

\par In Figure \ref{fig:r_pdr}, PDR plot of various network radius is shown. Number of GW is set to 1 and lowest SF assignment scheme is used. Increasing network radius, increases the number of under sensitivity transmissions thus decreases the PDR of network.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{r_pdr_g1_p0.01_s3600}.png}
\caption{PDR for various network radii.}
\label{fig:r_pdr}
\end{figure}

\par PDR plot of the lowest SF scheme and the smart prediction schemes are shown in Figure \ref{fig:prediction_pdr}. Randomly generated network radius is set to 5000 meters and number of GWs is set to 3. Prediction model needs nodes' locations and three GWs are enough to locate nodes positions by triangulation. In Table \ref{table:prediction_pdr}, PDR values for various network radii are presented. Both smart SVM and smart DTC schemes give better PDR than lowest SF schemes when number of nodes increases. Increasing number of nodes, increases number of interference. Smart schemes  improve network performance when LoRa interference is high. Moreover, smart schemes give better results when nodes are deployed closer to the GW, since nodes have margin to increase their SFs when they are deployed closer to the GW. If a node is far away from GW, then smart schemes cannot increase the SF to avoid interference since the asigned SF is already high.

\par In Table \ref{table:prediction_accuracy}, prediction accuracy of smart SVM and smart DTC schemes are presented for various network radii and number of nodes. Prediction accuracy is not directly proportional to network PDR. Correct prediction of a interfered transmission may not increase the PDR but increases the accuracy. Mart DTC gives better PDR results than smart SVM, even overall SVM prediction accuracy is high.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{prediction_pdr_r5000_g3_p0.01_s3600}.png}
\caption{PDR for lowest SF and smart SF schemes.}
\label{fig:prediction_pdr}
\end{figure}

\begin{table}
\centering
\caption{PDR for lowest SF and smart SF schemes.}
\label{table:prediction_pdr}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{\multirow{2}{*}{}}                            & \multicolumn{3}{c|}{\textbf{Number of Nodes}} \\ \cline{4-6}
\multicolumn{3}{|c|}{}                                             & 100           & 500           & 1000          \\ \hline
\multirow{12}{*}{\textbf{R (m)}} & \multirow{3}{*}{3000}  & Lowest & 97.8          & 86.0          & 72.3          \\ \cline{3-6}
                                 &                        & SVM    & 98.0          & 88.2          & 75.2          \\ \cline{3-6}
                                 &                        & DTC    & 97.8          & 89.8          & 78.7          \\ \cline{2-6}

@@

\par For space constraints, we omit results like network throughput, total transmit energy consumption. We invite readers to experiment the simulation tool with the parameters they like. \cite{simlorasf}.

\section{Conclusion} \label{Conclusion}
\par In this paper, we present a discrete event simulator which is developed from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. Moreover, we show how SF collisions can be avoided, hence, we propose a machine learning based solutions called smart DTC scheme and smart SVM scheme. We present simulation results for the lowest SF assignment scheme and the proposed schemes. Simulation results show that, the the proposed smart schemes can increase network performance for LoRaWAN networks, especially, when the nodes are deployed close to GWs.

\par As for future work, transmit power optimization can be included to the proposed smart schemes. In this paper, it is assumed that nodes always use maximum transmit power for uplink transmission, however nodes close to GWs can decrease transmit power to save energy. This will make transmissions more vulnerable to interference thus requires extra care. Moreover, other machine learning methods can be investigated for SF assignment enhancement. Reinforcement learning could be a good candidate. Also, other transmission parameters such as node id and transmission time can be included to the proposed scheme to the improve performance.
